Actuator line modeling of vertical-axis turbines

### 2018-05-04 reviews/reviewer1-response.md

# Response to reviewer 1

We thank the reviewer for his or her detailed comments, which we have addressed
to improve the manuscript significantly.

>L. 111. Can you define lambda0?

This definition has been added to the theory section of the manuscript.

>Fig. 4 and 5. Can you include a line with the experimental value of Cp to appreciate how the model converges to it?

This is a reasonable suggestion. However, we feel that this may confuse the
readers with the difference between checking the independence of the model
parameters with the validation reported later.

>L. 228. Is the mesh uniform in the three directions? It can be quite relevant to state how the vertical resolution of the turbine is.

Yes, the mesh was uniform in all directions. This detail has been added to the
manuscript.

>L. 233. Define zero-subscript quantities.

Thanks for catching this. Is it now noted that $\Delta t_0$ is the time step
chosed to run the final simulations.

>L. 268. The drag and thrust can be considered differently to CFTs. Although for the whole rotor the drag can be considered as the thrust, i.e. as typically in a cylinder, it can bring to confusion when drag coefficient calculations on a blade are performed, as these are dependent on x- and y- forces, and not just on the former.

This is a valid point. Both are defined in different coordinate systems.
The clarify, we have attempted to separate blade element
and rotor drag using lowercase and uppercase subscripts, respectively.

>Fig. 6. The results showed in this figure are very promising. The overestimation at high tip speed ratios have been commonly found in literature as the authors pointed out. However, the reviewer questions if such effects can arise from the use of k-epsilon turbulence model. This model is usually not considered when wall effects are relevant in the analysed flow. Thus, the k-omega model is more used as in the author’s work from reference 10. Why did you not consider k-omega which would allow to a closer comparison to those blade-resolved results? At high tip speed ratios, the blades might undergo quite light dynamic stall, and maybe your DS model is over-predicting this value and triggering the Cd and Cp curves. Nevertheless, this is an almost unavoidable consequence of actuator line models. The reviewer suggest that the turbulence model may contribute more than expected by the authors in such Cp prediction.

We have doubts that the turbulence model is causing much difference, since
near actuator line elements, there is not a wall model. There could be an
effect from a difference in turbulence production in the model from shear
caused by the momentum source. We hypothesize that these discrepancies arise
from secondary effects, or potentially deficiencies in the input foil data,
both of which affect the simpler momentum and vortex models.

>Fig. 11. The results of power coefficient from the second case show that the model is quite case dependent. The results show that the model is not able to represent the physics about this turbine and thus the actual accuracy of the ALM cannot be verified without any additional figure that shows how the flow field is. Can you provide transversal plane contours showing the actual flow field generated by the ALM? Could this inaccuracy be again due to the k-epsilon model?

We have added a snapshot of the internal flow field and force vectors to give
and idea of the level of detail, or lack thereof, regarding local flow near
the blades. Unfortunately there is no mechanism for the ALM to generate flow
features like the pair of counter rotating vortices induced by dynamic stall.
This is a compromise inherent to the ALM, which we have clarified. It is
advantageous when attempting to simulate array-scale flows, but not so much
for individual device optimization, despite our validation efforts focusing
on the latter.

>Section 9.4. The presented LES results are quite disappointing due to the large differences against the RANS results which were very good. The reviewer has few concerns about this section:
>
Can you specify more details about the number of processors used? It is true that LES demands more computational resources than RANS but with the meshes used for the ALM it should be affordable to run LES.

We ran the LES simulations on 4 processors, a note that has been added to the
paper. It was still relatively affordable compared to the blade-resolved RANS.
We have noted a 2 order of magnitude drop in computational cost.

>Overall LES results are poor. The predicted Cp dropped to 20% which seems to be derived from the resolution of the flow field instead of the ALM itself. While a good agreement is found for the velocity wake as seen in Fig. 13 and 15(left), the turbulent kinetic energy results show an unexpected over damping of the velocity fluctuations. The use of LES is herein motivated to achieve a better resolution of the turbulent fluctuations by explicitly resolving the large-scale flow structures, and in the present manuscript the TKE has dropped dramatically. Fig. 14 and 15 (right) show a lack of flow unsteadiness attributed by the authors to the influence of the SGS model. The spatial discretisation is not described and it may be possible that depending on it, the scheme introduced an extra damping typically used in implicit LES wherein no SGS is used. Nevertheless, the reviewer would recommend to double-check the kinetic energy results as is quite confusing that the velocity wake prediction is greatly improved while the velocity fluctuations are almost gone.

One important note is that the SGS TKE is not tracked by the Smagorinsky
model, so there will be a lack of energy there.

It is not likely that the numerical schemes are dissipating the energy, since
all 2nd order schemes are used.

TODO: Check on TKE calculation from LES model and finish this.

>For further validation, the authors may consider to plot RANS vs LES at closer distance to the turbine, e.g. x/D=0.5 and study how the turbulent kinetic energy decay is in both models, or a transversal plane showing the generation of the vortices.

### 2018-05-04 paper.tex
Start responses to reviewer 1

For a VAT or CFT reduced frequency can be reformulated in terms of the tip speed
ratio as \begin{equation} k = \frac{\lambda c}{2R}, \end{equation} which is then
also a function of solidity or chord-to-radius ratio $c/R$. As an example, a
large scale, relatively low solidity Darrieus wind turbine such as the Sandia 34
m diameter Test Bed, with an equatorial blade chord of 0.91 m \cite{Murray2011},
a reduced frequency $k=0.16$ is encountered based solely on angle of attack
oscillations at $\lambda=6$. For the two turbines studied here, reduced
frequencies at $\lambda_0$ are 0.27 and 0.17 (based on root chord length) for
the UNH-RVAT and RM2, respectively. Therefore, both cases will be inherently
unsteady, with these effects dominating the UNH-RVAT's behavior. Hence, an
accurate unsteady aerodynamics model is important to predicting blade loading.

@@

Verification for sensitivity to spatial and temporal grid resolution was
performed for both the UNH-RVAT and RM2 RANS cases at their optimal tip speed
rations, the results from which are plotted in
Figure~\ref{fig:RVAT-ALM-verification} and
Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
strategy employed in \cite{Bachant2016-BR-CFD}, the mesh topology was kept
constant, and the resolution was scaled proportional to the number of cells in
the $x$-direction $N_x$ for the base hexahedral mesh. Both models displayed low
sensitivity to the number of time steps per revolution. Spatial grid dependence,
however, was more important.

Final spatial grid resolutions were chosen as $N_x=48$ for both the UNH-RVAT and
RM2 cases. Time steps were chosen as $\Delta t = 0.01$ and $\Delta t = 0.005$
seconds for the UNH-RVAT and RM2 respectively, which correspond to approximately
200 steps per revolution. The chosen values should provide $C_P$ predictions
within one percentage point of the true solution, which is on the order of the
expanded uncertainty of the experimental measurements. Note that for computing
performance curves, the number of steps per revolution was kept constant, i.e.,
the time step was adjusted to $\Delta t = \Delta t_0 \lambda_0 / \lambda$.

@@

@@

Since the computational cost of LES is significantly higher than RANS,
verification with respect to grid dependence was not performed. Instead, mesh
resolution was chosen relative to similar studies of turbine wake ALM LES. Of
the studies surveyed
\cite{Shamsoddin2014,Archer2013,Martinez-Tossas2015a,Troldborg2007}, the mesh
resolution ranged from 18--64 points per turbine diameter. The mesh here was set
accordingly by using a 16 point per meter base mesh, and refining twice in a
region containing the turbine to produce a 64 point per turbine diameter/height
resolution. The solver was run with a 0.002 second time step, which is
significantly within the limit described by \cite{Martinez-Tossas2015}, where an
actuator line element may not pass through more than one cell per time step.
With these resolutions computation times were $O(10)$ CPU hours per second of
simulated time, which is approximately two orders of magnitude lower than for
blade-resolved RANS.

>>>>

For a VAT or CFT reduced frequency can be reformulated in terms of the tip speed
ratio as \begin{equation} k = \frac{\lambda c}{2R}, \end{equation} which is then
also a function of solidity or chord-to-radius ratio $c/R$. As an example, a
large scale, relatively low solidity Darrieus wind turbine such as the Sandia 34
m diameter Test Bed, with an equatorial blade chord of 0.91 m \cite{Murray2011},
a reduced frequency $k=0.16$ is encountered based solely on angle of attack
oscillations at $\lambda=6$. For the two turbines studied here, reduced
frequencies at the tip speed ratio of peak power coefficient, $\lambda_0$,
are 0.27 and 0.17 (based on root chord length) for
the UNH-RVAT and RM2, respectively. Therefore, both cases will be inherently
unsteady, with these effects dominating the UNH-RVAT's behavior. Hence, an
accurate unsteady aerodynamics model is important to predicting blade loading.

@@

Verification for sensitivity to spatial and temporal grid resolution was
performed for both the UNH-RVAT and RM2 RANS cases at their optimal tip speed
rations, the results from which are plotted in
Figure~\ref{fig:RVAT-ALM-verification} and
Figure~\ref{fig:RM2-ALM-verification}, respectively. Similar to the verification
strategy employed in \cite{Bachant2016-BR-CFD}, the mesh topology was kept
constant, and the resolution was scaled proportional to the number of cells in
the $x$-direction $N_x$ for the base hexahedral mesh, which was essentially
uniform in resolution in all directions. Both models displayed low
sensitivity to the number of time steps per revolution. Spatial grid dependence,
however, was more important.

Final spatial grid resolutions were chosen as $N_x=48$ for both the UNH-RVAT and
RM2 cases. Time steps were chosen as $\Delta t_0 = 0.01$ and
$\Delta t_0 = 0.005$
seconds for the UNH-RVAT and RM2 respectively, which correspond to approximately
200 steps per revolution. The chosen values should provide $C_P$ predictions
within one percentage point of the true solution, which is on the order of the
expanded uncertainty of the experimental measurements. Note that for computing
performance curves, the number of steps per revolution was kept constant, i.e.,
the time step was adjusted to $\Delta t = \Delta t_0 \lambda_0 / \lambda$,
where $\lambda_0$ was the tip speed ratio roughly corresponding to peak $C_P$
as observed in experiments, which was used for verifying the numerical
sensitivity to spatial and temporal resolution.

@@

The internal rotor flow about the horizontal center plane, along with
the directions and relative magnitudes of the actuator line force vectors
are shown in Figure~\ref{fig:RM2-internal}.
It can be seen how flow features like dynamic stall vortices are not
captured with the ALM, despite forces themselves being modulated by the
dynamic stall model. This combined with the diffusive nature of the
RANS model helps explain discrepancies in predicted power coefficient.
With the computational affordability of the ALM coupled with RANS comes a
compromise in flow detail, which should be acceptable for array-level
simulation, but for individual device design a blade-resolved
simulation would be more advisable.

    \includegraphics[width=0.8\textwidth]{rm2-vel-force-6-seconds}

    \caption{Velocity magnitude and actuator line force field vectors for the
    RM2 RANS simulation at 6 seconds simulated time.}

    \label{fig:RM2-internal}
\end{figure}

@@

Since the computational cost of LES is significantly higher than RANS,
verification with respect to grid dependence was not performed. Instead, mesh
resolution was chosen relative to similar studies of turbine wake ALM LES. Of
the studies surveyed
\cite{Shamsoddin2014,Archer2013,Martinez-Tossas2015a,Troldborg2007}, the mesh
resolution ranged from 18--64 points per turbine diameter. The mesh here was set
accordingly by using a 16 point per meter base mesh, and refining twice in a
region containing the turbine to produce a 64 point per turbine diameter/height
resolution. The solver was run with a 0.002 second time step, which is
significantly within the limit described by \cite{Martinez-Tossas2015}, where an
actuator line element may not pass through more than one cell per time step.
With these resolutions computation times were $O(10)$ CPU hours per second of
simulated time (on 4 processors),
which was approximately two orders of magnitude lower than for
blade-resolved RANS.
